Does It Scale? Symbolic Execution of PHP Web Applications

### 2017-01-24 paper.tex
language refactoring

\documentclass[sigconf, preprint]{acmart}

@@

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

@@

%\author{Stefan Mühlbauer}
\affiliation{%
  %\institution{Technische Universität Braunschweig}
  %\streetaddress{P.O. Box 1212}
  %\city{Braunschweig} 
  %\state{Germany} 
  %\postcode{43017-6221}
}
%\email{s.muehlbauer@tu-bs.de}

@@

\begin{abstract}
Dynamic web applications have become widely popular and are to a large portion
implemented in the scripting language PHP. Dynamic languages like PHP though
challenge static analysis, as they generate code at run-time. One approach to unfold their
staged nature and to approximate client page output is symbolic execution.
Knowledge about possible output variants enables a range of developer tools as
well as analyses for vulnerability detection. However, previous tools have only
been evaluated for smaller systems and scalability for larger systems has not
been assessed.

@@

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
\begin{CCSXML}
<ccs2012>
 <concept>
  <concept_id>10010520.10010553.10010562</concept_id>
  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
  <concept_significance>500</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010575.10010755</concept_id>
  <concept_desc>Computer systems organization~Redundancy</concept_desc>
  <concept_significance>300</concept_significance>
 </concept>
 <concept>
  <concept_id>10010520.10010553.10010554</concept_id>
  <concept_desc>Computer systems organization~Robotics</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
 <concept>
  <concept_id>10003033.10003083.10003095</concept_id>
  <concept_desc>Networks~Network reliability</concept_desc>
  <concept_significance>100</concept_significance>
 </concept>
</ccs2012>  
\end{CCSXML}

\ccsdesc[500]{Computer systems organization~Embedded systems}
\ccsdesc[300]{Computer systems organization~Redundancy}
\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

@@

\keywords{ACM proceedings, \LaTeX, text tagging}

@@

A common characteristic of all dynamic web applications is their
\emph{staged nature}: A dynamic web application as a whole
consists of both static code, such as scripts, and dynamically generated code,
such as client page output. The latter code, although assembled at
run-time, may contain client-side parts of the web application, such as
JavaScript. Therefore, to study dynamic web applications in its entirety, we
need to consider both static as well as dynamic aspects of those systems.

@@

To investigate the question, whether we can have a practical and scalable
symbolic execution engine, we re-implemented the symbolic engine with the
specifications of the previous symbolic execution semantics
for \sf{PhpSync} \cite{Nguyen:2014:BCG:2635868.2635928} for PHP and
extended it with additional features, such as object-oriented programming. We
evaluate our symbolic execution engine for several large-scale and modern PHP systems.

During the introduction of new semantics for object-oriented programming we
faced a trade-off between accuracy and scalability: Whilst exploring multiple
paths, for method invocations the number of target objects may grow rapidly and
infeasible. Moreover, based on observations we identify limitations of symbolic execution
for PHP. Dynamic features such as dynamic function invocations and dynamic
include for an accurate execution require precisely resolved include targets
and function names respectively. Since these expressions  can contain symbolic
values at run-time, for many cases this restricts a symbolic execution engine
from further execution as callable code sections become inaccessible.
Although some effort can be spent to statically resolve dynamic include
expressions \cite{hills2014static}, or to simply provide static counterparts manually, resolution
is often not trivial since expressions can contain symbolic values assumed instead
of concrete program input, including databases, user inputs or configuration
files. We empirically evaluate our symbolic execution engine for a corpus
including large and actively maintained PHP systems.

@@

\section{State Of the Art}\label{sec:stateoftheart}
This Section recaps the idea of symbolic
execution, the tools using output approximation based on it and whose evaluation
we criticize, and current conceptual and language-related limitations.

@@

Therefore, for our execution engine to be sound, we were bound to execute
invocated methods for multiple targets since we cannot assert that all values
contained in the value summary are of the same type. Nevertheless, to scale the
target object space, in addition, we implemented execution modes only allowing
execution of a method for a subset of possible targets and assuming symbolic
return values for the remaining targets. This execution mode was necessary for
\sf{WordPress} for several, but not all entry points that we studied, in order
to terminate. All subsequent analyses in this paper do not use this execution
mode, yet it is provided in our implementation.

\subsubsection{Dynamic Language Features}
\label{sec:experiment_dynamicfeatures} 
Aside from the previous issues, we identified two dynamic language features
that we were not able to execute accurately. Dynamic features in PHP are
features that take as input rather a expression that is resolved at run-time
than a static constant value. For \sf{WordPress}, we faced missing files and
functions that are included using dynamic includes and dynamic function
invocation respectively.

First, file inclusion in PHP usually is managed by statements like \tt{include},
\tt{include\_once} or \tt{require} and \tt{require\_once}. For transitive
includes, statements ending with \tt{\_once} will not include the same file
twice. The include statements make an attempt to include a file, but do not
return a warning in contrast to the \tt{require} statements. Each of these
statements takes as input either a string constant (static include) or an
expression that is evaluated at run-time (dynamic include). The string that the
expression evaluates to represents an absolute or relative paths to the target
file that the interpreter finally attempts to include.

@@

\section{Evaluation} \label{sec:evaluation}
As already stated in the previous Section \ref{sec:experience_report}, we
encountered several problems during the implementation of our symbolic
execution engine for \sf{WordPress}. The main items of critique of the symbolic
execution engine and tools described in previous work
\cite{Nguyen:2011:AFH:2190078.2190142,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140}
are the missing support for widely-used language features, and the evaluation
only using small-scale systems. Although evaluating a tool for a small systems
can be a valuable proof-of-concept, it does not necessarily cover the question
of whether a tool is practical on a larger-scale. Therefore in the following we
present our methods to evaluate practicality of our symbolic output
approximation. Moreover, we investigate to which extent the limitations encountered
with \sf{WordPress} apply for a wider choice of larger PHP systems.

@@

Second, we ask: \textit{Do system size and usage of modern language features
factors affect the practicality of our symbolic execution engine?} We
selected both a range of large-scale and small-scale modern PHP systems to
better separate and understand the circumstances under which limitations might affect practicality. For the selection of large-scale systems
we borrow three systems from a case study by Hills et al. \cite{hills2015variable} addressing the
feature usage patterns and frequencies in PHP systems. Some more detailed description of this
case study can be found in Section \ref{sec:related_work}. For our choice of modern small-scale systems we
selected five systems from a list of recent Content Management Systems (CMS) written in PHP \cite{codegeekz}. For both
selections we were bound to select a smaller number of systems than we intended
to since the parser used in our implementation,
\sf{Quercus}\footnote{\sf{Quercus} is an open-source Java implementation of PHP
5 by Caucho Technology. We reused the parser front-end for our implementation.
\sf{Quercus} can be obtained from \url{http://quercus.caucho.com}.}, did not
support a number of language features.
The full list of PHP systems for that we evaluate our symbolic execution engine
along with descriptive code metrics can be found in Table \ref{corpus}.

@@

\subsubsection{Where did we miss output?}\label{WhatLiteralsDidWeMiss?}
Our initial approach to understand missed output candidates is to find out
whether their surrounding program code cannot not accessed, and if
so, why. We are interested in those output candidates that are located in HTML
files (possibly with nested PHP), or located in a function definition. Plain
HTML files do not require any further execution and represent output candidates
that just need to be included properly to be part of the output. Aside, output
candidates that are part of a function definition are only missed if the
corresponding function is never invocated at run-time, or simply dead code.

We started classifying missed output candidates by their string literal
context. As the classification statistics in Figure \ref{coverage} illustrate,
for almost all systems tested (except for \textsf{AddressBook} and
\textsf{TimeClock}) not-reached output candidates have a function context.

@@

Aside from being limited by dynamic features, we exercised replacing dynamic
features with respective static counterparts in order to bypass limitations by
manually providing expressions that can be resolved precisely.To use
approaches incorporating symbolic execution with dynamic web applications in
the long run, more concrete input, similar to test oracles, needs
to be provided to decrease the number of assumed symbolic inputs and avoid
dynamic features being confronted with symbolic values. Existing work approaching
this goal \cite{artzi_finding_2008,artzi_finding_2010,DynamicWassermann}
incorporates dynamic symbolic execution, i.e. concrete execution alongside symbolic execution.

Also, if the usage of tools using static output approximation with symbolic
execution is intended, avoiding dynamic features that are prone to be inaccurately executed for
symbolic execution  may be an adjustment strategy to program in an analyzable
way.

@@

\paragraph{The road so far.}
For this paper, the extension of the symbolic execution \sf{PhpSync} \cite{Nguyen:2014:BCG:2635868.2635928}
support for object-oriented programming was not the main goal, yet necessary to
symbolically execute modern PHP systems. The observations regarding object
target space explosion for method calls documented in Section
\ref{sec:experiment_dynamicfeatures} are only anecdotal and, with the exception
of \sf{WordPress}, we did not notice any significant impact on scalability.
Nevertheless, more research is required to better and systematically understand
possible drawbacks of object-oriented symbolic execution. 
This question though is beyond the scope of this paper.
Future work might address design guidelines to enable programming analyzable applications, tool support to scale
the number of assumed symbolic inputs or systematically provide concrete data
to increase the practical benefit of the tools described above.

>>>>

\documentclass[sigconf]{acmart}

% removes copyright & conference info
\settopmatter{printacmref=false} % Removes citation information below abstract
\renewcommand\footnotetextcopyrightpermission[1]{} % removes footnote with  conference information in first column 
\pagestyle{plain} % removes running headers

@@

% Copyright
%\setcopyright{none}
%\setcopyright{acmcopyright}
%\setcopyright{acmlicensed}
%\setcopyright{rightsretained}
%\setcopyright{usgov}
%\setcopyright{usgovmixed}
%\setcopyright{cagov}
%\setcopyright{cagovmixed}

@@

\author{Stefan Mühlbauer}
\affiliation{%
  \institution{Technische Universität Braunschweig}
  %\streetaddress{P.O. Box 1212}
  %\city{Braunschweig} 
  %\state{Germany} 
  %\postcode{43017-6221}
}
%\email{s.muehlbauer@tu-bs.de}

@@

\begin{abstract}
Dynamic web applications have become widely popular and are to a large portion
implemented in the scripting language PHP. Dynamic languages like PHP though
challenge static analysis, as they generate code at run-time. One approach to unfold their
staged nature and to approximate client page output is symbolic execution.
Knowledge about possible output variants enables a range of developer tools as
well as analyses for vulnerability detection. However, previous tools that
leverage symbolically approximated output have only been evaluated for smaller
systems and scalability for larger systems has not been assessed.

@@

%
% The code below should be generated by the tool at
% http://dl.acm.org/ccs.cfm
% Please copy and paste the code instead of the example below. 
%
%\begin{CCSXML}
%<ccs2012>
% <concept>
%  <concept_id>10010520.10010553.10010562</concept_id>
%  <concept_desc>Computer systems organization~Embedded systems</concept_desc>
%  <concept_significance>500</concept_significance>
% </concept>
% <concept>
%  <concept_id>10010520.10010575.10010755</concept_id>
%  <concept_desc>Computer systems organization~Redundancy</concept_desc>
%  <concept_significance>300</concept_significance>
% </concept>
% <concept>
%  <concept_id>10010520.10010553.10010554</concept_id>
%  <concept_desc>Computer systems organization~Robotics</concept_desc>
%  <concept_significance>100</concept_significance>
% </concept>
% <concept>
%  <concept_id>10003033.10003083.10003095</concept_id>
%  <concept_desc>Networks~Network reliability</concept_desc>
%  <concept_significance>100</concept_significance>
% </concept>
%</ccs2012>  
%\end{CCSXML}

%\ccsdesc[500]{Computer systems organization~Embedded systems}
%\ccsdesc[300]{Computer systems organization~Redundancy}
%\ccsdesc{Computer systems organization~Robotics}
%\ccsdesc[100]{Networks~Network reliability}

@@

%\keywords{ACM proceedings, \LaTeX, text tagging}

@@

A common characteristic of all dynamic web applications is their
\emph{staged nature}: A dynamic web application as a whole
consists of both static code, such as scripts, and dynamically generated code,
such as client page output. The latter code, although generated at
run-time, may contain client-side parts of the web application, such as
JavaScript. Therefore, to study dynamic web applications in its entirety, we
need to consider both static as well as dynamic aspects of those systems.

@@

To investigate the question, whether we can have a practical and scalable
symbolic execution engine, we re-implement the symbolic engine with the
specifications of the previous symbolic execution semantics
for \sf{PhpSync} \cite{Nguyen:2014:BCG:2635868.2635928} for PHP and
extend it with additional features, such as object-oriented programming. We
evaluate our symbolic execution engine for several large-scale and modern PHP systems.

During the introduction of new semantics for object-oriented programming we
face a trade-off between accuracy and scalability: Whilst exploring multiple
paths, for method invocations the number of target objects can grow rapidly and
become infeasible. Moreover, based on observations we identify limitations of
symbolic execution for PHP. For an accurate execution dynamic features, such as
dynamic function invocations and dynamic includes, require precisely resolved
include targets and function names respectively. Since these expressions  can
contain symbolic values at run-time, for many cases this restricts a symbolic
execution engine from further execution as callable code sections become
inaccessible. 
Although some effort can be spent to statically resolve dynamic include
expressions \cite{hills2014static}, or to simply provide static counterparts manually, resolution
is often not trivial since expressions can contain symbolic values assumed instead
of concrete program input, including databases, user inputs or configuration
files. We empirically evaluate our symbolic execution engine for a corpus
including large and actively maintained PHP systems.

@@

\section{State Of the Art}\label{sec:stateoftheart}
This Section recaps the idea of symbolic
execution, the tools using output approximation based on it and whose evaluation
we criticize as well as current conceptual and language-related limitations.

@@

Therefore, for our execution engine to be sound, we are bound to execute
invocated methods for multiple targets since we cannot assert that all values
contained in the value summary are of the same type. Nevertheless, to scale the
target object space, in addition, we implemented execution modes only allowing
execution of a method for a subset of possible targets and assuming symbolic
return values for the remaining targets. This execution mode was necessary for
\sf{WordPress} for several, but not all entry points that we studied, in order
to terminate. All subsequent analyses in this paper do not use this execution
mode, yet it is provided in our implementation.

\subsubsection{Dynamic Language Features}
\label{sec:experiment_dynamicfeatures} 
Aside from the previous issues, we identified two dynamic language features
that, in many cases, we were not able to execute accurately. Dynamic features in
PHP are features that take as input rather a expression that is resolved at run-time
than a static constant value. For \sf{WordPress}, we faced missing files and
functions that are included using dynamic includes and dynamic function
invocation respectively.

First, file inclusion in PHP usually is managed by statements like \tt{include},
\tt{include\_once} or \tt{require} and \tt{require\_once}. For transitive
includes, statements ending with \tt{\_once} will not include the same file
twice. The include statements make an attempt to include a file, but do not
return a warning in contrast to the \tt{require} statements. Each of these
statements takes as input either a string constant (static include) or an
expression that is evaluated at run-time (dynamic include). The string that the
expression evaluates to represents an absolute or relative path to the target
file that the interpreter finally attempts to include.

@@

\section{Evaluation} \label{sec:evaluation}
As already stated in the previous Section \ref{sec:experience_report}, we
encountered several problems during the implementation of our symbolic
execution engine for \sf{WordPress}. The main items of critique of the symbolic
execution engine and tools described in previous work
\cite{Nguyen:2011:AFH:2190078.2190142,Nguyen:2014:BCG:2635868.2635928,Nguyen:2015:CPS:2786805.2786872,Nguyen:2015:VIS:2819009.2819140}
are the missing support for widely-used language features, and the evaluation
only using small-scale systems. Although evaluating a tool for a small systems
can be a valuable proof of concept, it does not necessarily cover the question
of whether a tool is practical on a larger-scale. Therefore in the following we
present our methods to evaluate practicality of our symbolic output
approximation. Moreover, we investigate to which extent the limitations encountered
with \sf{WordPress} apply for a wider choice of larger PHP systems.

@@

Second, we ask: \textit{Do system size and usage of modern language features
factors affect the practicality of our symbolic execution engine?} We
selected both a range of large-scale and small-scale modern PHP systems to
better separate and understand the circumstances under which limitations might affect practicality. For the selection of large-scale systems
we borrowed three systems from a case study by Hills et al.
\cite{hills2015variable} addressing the feature usage patterns and frequencies in PHP systems. Some more detailed description of this
case study can be found in Section \ref{sec:related_work}. For our choice of modern small-scale systems we
selected five systems from a list of recent Content Management Systems (CMS) written in PHP \cite{codegeekz}. For both
selections we were bound to select a smaller number of systems than we intended
to since the parser used in our implementation,
\sf{Quercus}\footnote{\sf{Quercus} is an open-source Java implementation of PHP
5 by Caucho Technology. We reused the parser front-end for our implementation.
\sf{Quercus} can be obtained from \url{http://quercus.caucho.com}.}, did not
support a number of language features.
The full list of PHP systems for that we evaluate our symbolic execution engine
along with descriptive code metrics can be found in Table \ref{corpus}.

@@

\subsubsection{Where did we miss output?}\label{WhatLiteralsDidWeMiss?}
Our initial approach to understand missed output candidates is to find out
whether their surrounding program code cannot not accessed, and if
so, why. We are interested in those output candidates that are located in HTML
files (possibly with nested PHP), or located in a function definition. Plain
HTML files do not require any further execution and represent output candidates
that just need to be included properly to be part of the output. Aside, output
candidates that are part of a function definition are only missed if the
corresponding function is never invocated at run-time or simply just dead code.

We started classifying missed output candidates by their string literal
context. As the classification statistics in Figure \ref{coverage} illustrate,
for almost all systems tested (except for \textsf{AddressBook} and
\textsf{TimeClock}) not-reached output candidates had a function context.

@@

Aside from being limited by dynamic features, we exercised replacing dynamic
features with respective static counterparts in order to bypass limitations by
manually providing expressions that can be resolved precisely.To use
approaches incorporating symbolic execution with dynamic web applications in
the long run, more concrete input, similar to test oracles, needs
to be provided to decrease the number of assumed symbolic inputs and avoid
dynamic features being confronted with symbolic values. Existing work approaching
this goal \cite{artzi_finding_2008,artzi_finding_2010,DynamicWassermann}
incorporates dynamic symbolic execution, i.e. concrete execution alongside symbolic execution.
Also, if the usage of tools using static output approximation with symbolic
execution is intended, avoiding dynamic features that are prone to be inaccurately executed for
symbolic execution  may be an adjustment strategy to program in an analyzable
way.

@@

\paragraph{The road so far.}
For this paper, the extension of the symbolic execution \sf{PhpSync} \cite{Nguyen:2014:BCG:2635868.2635928}
support for object-oriented programming was not the main goal, yet necessary to
symbolically execute modern PHP systems. The observations regarding object
target space explosion for method calls documented in Section
\ref{sec:experiment_dynamicfeatures} are only anecdotal and, with the exception
of \sf{WordPress}, we did not notice any significant impact on scalability.
Nevertheless, more research is required to better and systematically understand
possible drawbacks of object-oriented symbolic execution. 
This question though is beyond the scope of this paper.
Future work might address design guidelines to enable programming analyzable applications, tool support to scale
the number of assumed symbolic inputs or systematically provide concrete data
to increase the practicality of the tools described above.
